Finite Complete Suites for CSP Refinement Testing

### 2019-04-16 scp-journal/article-revision-1/main.tex
ana reviewer

A prefixing operator $e \then P$ defines a process that is ready to engage in
the event $e$, pending agreement of its environment to synchronise. After $e$
occurs, the process behaves as defined by $P$. The environment can be other
processes, in parallel, or the environment of a system as a whole.

Two forms of choice support branching behaviour. An external choice $P
\extchoice Q$ between processes $P$ and $Q$ offers to the environment the
initial events of $P$ and $Q$. Once a synchronisation takes place, the
process that has offered the event that has occurred is chosen and the other
is discarded. In an internal choice $P \intchoice Q$, the environment does
not have an opportunity to interfere:~the choice is made by the process.
%
\begin{example}\label{example:CSP}
  We consider the processes $P$, $Q$, and $R$ defined below. $P$ is initially
  ready to engage in the event $a$, and then makes an internal choice to
  behave like either $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators define a choice over a collection of
processes $P(i)$. If the index set $I$ is empty, the external choice is the
process $\Stop$, which deadlocks:~does not engage into any event or
terminate. For an external choice $\Extchoice e: A @ e \then P(e), A\subseteq
\Sigma$ over a set $A$ of events, we use the abbreviation $e:A \then P(e)$.
An iterated internal choice is not defined for an empty index set.

The branches of an external choice can be protected by \emph{guards}. The
 process
$$
P(x_1,\dots,x_n) = \text{bexpr}_1(x1,\dots,x_n) \& Q(x_1,\dots,x_n) \extchoice 
\text{bexpr}_2(x1,\dots,x_n) \& R(x_1,\dots,x_n)
$$
parametrised over $x_1,\dots,x_n$ branches according to Boolean expressions
$\text{bexpr}_1(x1,\dots,x_n)$ and $\text{bexpr}_2(x1,\dots,x_n)$ over 
variables from $\{ x_1,\dots,x_n\}$: if $\text{bexpr}_1(x1,\dots,x_n)$ evaluates to
$\ist$ and $\text{bexpr}_2(x1,\dots,x_n)$ to $\isf$, $P$ behaves like $Q$. 
If $\text{bexpr}_1(x1,\dots,x_n)$ evaluates to
$\isf$ and $\text{bexpr}_2(x1,\dots,x_n)$ to $\ist$, $P$ behaves like $R$. If both
Boolean expressions evaluate to $\ist$, $P$ behaves as $Q\extchoice R$. Finally, if
both Boolean expressions evaluate to $\isf$, $P$ behaves like $\Stop$.

@@

Semantic definitions specify, for each operator, how the traces or failures
of the resulting process can be calculated from those of each operand. For
example, for internal choice, $\fails(P\intchoice Q) =
\fails(P)\cup\fails(Q)$; see~\cite[p.~210]{Roscoe:1997:TPC:550448} for a
comprehensive list of these definitions covering $\trc(P)$ and $\fails(P)$.

@@

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined by $\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed,
$P \lessdet_F Q$ implies $(s,\varnothing)\in\fails(P)$ for all traces
$s\in\trc(Q)$. So, for divergence-free processes, failures refinement implies
traces refinement. Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$ as proven
in~\cite{DBLP:conf/icfem/CavalcantiG07}.
%
\begin{equation}%\label{thm:fref-tref-conf}
\label{eq:failconf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite processes, since refusals are subset-closed, $\refs(P/s)$ can be
constructed from the set of \emph{maximal refusals}.
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$. % as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
defined by $\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. Nondeterministic behaviour in a given process
state is reflected by non-empty intersections between initials and maximal refusals. This 
is illustrated by the following example.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$. Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

@@

% =========================================================================
\paragraph{Graph Products} \label{sec:GP} For proving our main theorems, it is
necessary to consider the \emph{product} of normalised transition graphs. We
need this only for the investigation of corresponding traces in reference
processes and processes for SUTs. So, the labelling of nodes with maximal
refusals or minimal acceptances is disregarded in the product construction.
We consider two normalised transition graphs
\[
G_i = ( N_i, \ii n_i, \Sigma, t_i : N_i\times\Sigma \pfun N_i, r_i : N_i \fun \mathbb{P}\mathbb{P}(\Sigma)),\qquad i = 1,2,
\]
over the same alphabet $\Sigma$. Their product is defined by
%
\begin{eqnarray}
G_1\times G_2 & = & (N_1\times N_2,(\ii n_1,\ii n_2), t:(N_1\times N_2)\times\Sigma\pfun (N_1\times N_2))
\\
\dom~t & = & \{ ((n_1,n_2),e)\in (N_1\times N_2)\times\Sigma | (n_1,e)\in\dom~t_1\wedge (n_2,e) \in\dom~t_2 \}
\\
t((n_1,n_2),e) & = & (t_1(n_1,e),t_2(n_2,e))\ \text{for $((n_1,n_2),e)\in\dom~t$}
\end{eqnarray}
%
The following lemma is used in the proof of our main theorem.
%
\begin{lemma}\label{lemma:reachproduc}
If $G_1$ has $p$ states and $G_2$ has $q$ states, then every reachable state
$(n_1,n_2)$ of the product graph $G_1\times G_2$ can be reached by a trace
%%%$s\in\trc(G_1)\cap \trc(G_2)$  we don't need this
of maximal length $(pq-1)$.
\end{lemma}
\begin{proof}
The product graph $G_1\times G_2$ has at most $pq$ states. The empty trace $\varepsilon$
reaches its initial state $(\ii n_1,\ii n_2)$. Applying Lemma~\ref{lemma:extendV}
$(pq-1)$ times with $V=\{\varepsilon \}$ implies that $G_1\times G_2$ reaches
all of its reachable states (there are at most $pq$ of them) under
$V' = V \cup V.\Sigma\cup\dots \cup V.\Sigma^{(pq-1)}$. The maximal length of traces in
$V'$ is $(pq-1)$.
\xbox
\end{proof}
%
This concludes our presentation of CSP and of results regarding its
semantics that are used in the next section.

@@

\paragraph{Definition} The main idea of the underlying test strategy for failures
refinement is based on solving a \emph{hitting set problem}. Given a finite
collection of finite sets ${\cal C} = \{ A_1,\dots,A_n\}$, such that each
$A_i$ is a subset of a universe $\Sigma$, a \emph{hitting set}
$H\subseteq\Sigma$ is a set satisfying the following property.
%
\begin{equation}
  \label{eq:hit}
  \forall A\in {\cal C}: H\cap A \neq\varnothing.
\end{equation}
%
A \emph{minimal hitting set} is a hitting set that cannot be further reduced
without losing the characteristic property (\ref{eq:hit}). By $\minhits({\cal
C})$ we denote the collection of minimal hitting sets for a collection ${\cal
C}$. For the pathological case where $C$ contains an empty set,
$\minhits({\cal C})$ is also empty. The problem of determining minimal hitting sets is %%% \cite{Book1975-BOOKRM,5533149}
NP-hard~\cite{5533149}. We see below, however, that using minimal hitting
sets, we can reduce the effort of testing for failures refinement from a
factor of $2^{|\Sigma|}$ to a factor that equals the number of minimal
hitting sets.

% -----------------------------------------------------------------------------
\paragraph{Minimal Hitting Sets of Minimal Acceptances} In this article, we are
interested in the minimal hitting sets of minimal acceptances; for these, the
abbreviated notation
$%\[ \
\minhits(P/s) = \minhits(\minaccs(P/s))
$ %\]
is used. The minimal hitting sets
of minimal acceptances may be alternatively characterised by means of the
failures of a process as is done in~\cite{DBLP:conf/icfem/CavalcantiG07}. To
this end, in~\cite{DBLP:conf/icfem/CavalcantiG07}, the authors define, for
any collection ${\cal C}\subseteq 2^\Sigma$ of subsets from $\Sigma$
\begin{equation}\label{eq:minsub}
\minsub({\cal C}) = \{ A\in {\cal C}~|~\forall B\in {\cal C}:
B\subseteq A\Rightarrow B = A \}.
\end{equation}
The collection $\minsub({\cal C})$ contains all those sets of
${\cal C}$ that are not true supersets of other members of ${\cal C}$. With
this definition, the relation between failures and minimal hitting sets of
minimal acceptances is established in the following lemma.
%
\begin{lemma}
\label{lemma:failureshittingsets}
For any trace $s$ of CSP process $P$, define
$%\[
{\cal A}_s = \{ A\subseteq \Sigma~|~(s,A)\not\in\fails(P)  \}.
$ %\]
Then
$ %\[
\minhits(P/s) = \minsub({\cal A}_s)
$ %\]
for all traces $s$ of $P$.
\end{lemma}
\begin{proof}
We derive
\begin{argue}
(s,A)\not\in \fails(P)
\\
\Leftrightarrow A\not\in Ref(P/s)
\\
\Leftrightarrow  \forall R\in maxRef(P/s): A\not\subseteq R
\\
\Leftrightarrow   \forall B\in minAcc(P/s): A\not\subseteq (\Sigma -
B)& this follows from (\ref{eq:accref})
\\
\Leftrightarrow  \forall B\in minAcc(P/s): A\cap B\neq \varnothing
\\
\Leftrightarrow  A \text{ is a (not necessarily minimal)}
              \text{ hitting set of } minAcc(P/s)
\end{argue}
%
This derivation is valid  for arbitrary $A$ satisfying $(s,A)\not\in {\text{failures}}(P)$.
Specialising it to {\it minimal} sets $A$ satisfying  $(s,A)\not\in {\text{failures}}(P)$
proves the statement of the lemma.
\xbox
\end{proof}

@@

To prove ``$(2) \Rightarrow (1)$'', we assume that (2) holds, but that $P\
conf\ Q$ does {\it not} hold. According to Lemma~\ref{lemma:tgtrcref},
(\ref{eq:failrefb}), there exists $s\in\trc(P)\cap \trc(Q)$ such that
\[
\exists A_Q\in\minaccs(G(Q)/s): \forall A_P\in\minaccs(G(P)/s):
A_P\not\subseteq A_Q \qquad (*)
\]
Let $A$ be such an acceptance set $A_Q$ fulfilling (*).
Define
$%\[
\overline H = \bigcup\{ A_P \setminus A~|~A_P\in\minaccs(G(P)/s) \}.
$ %\]
Since $A_P \setminus A \neq\varnothing$ for all $A_P$ because of (*),
$\overline H$ is a hitting set of $\minaccs(G(P)/s)$ which has an  empty
intersection with $A$. Minimising $\overline H$ yields   a minimal hitting
set $H\in \minhits(P/s)$ which is {\it not} a hitting set of
$\minaccs(G(Q)/s)$, a contradiction to Assumption~2.
This completes the proof of the lemma.
\xbox
\end{proof}
%
We note that $\minaccs(P) = \{ \varnothing \}$ if $P = Q\intchoice \Stop$.
Since $\Stop$ accepts nothing, its minimal acceptance is $\varnothing$, and
this carries over to $Q\intchoice \Stop$.  From (\ref{eq:failrefb}) we
conclude that $\varnothing\in\minaccs(P)$ implies $\minaccs(P) = \{
\varnothing \}$. This clarifies that $\minhits(P/s)$ is empty if, and only
if, $\minaccs(P) = \{ \varnothing \}$. The proof of Lemma~\ref{lemma:hseta}
covers the situations where $\minaccs(P/s) = \{ \varnothing \}$ and so
$\minhits(P/s) = \varnothing$. Trivially,
\begin{equation}
\label{eq:minhitminaccempty}
\minaccs(P/s) = \{ \varnothing \} \iff \minhits(P/s) = \varnothing
\end{equation}
holds.

@@

It is assumed that the events $\efail$ and $\epass$, indicating verdicts FAIL
and PASS for the test execution, are not included in $\Sigma$. Since we
assume that $Q$ is free of livelocks, it is guaranteed that events $\efail$
or $\epass$ always become visible, if they are the only events $U_F(j)/s$ is
ready to engage in: if $U_F(j)/s$ can only produce $\epass$ or $\efail$, the
occurrence of these events can never be blocked due to a livelock, since the only
  process executing concurrently with  $U_F(j)$  is livelock-free  $Q$.

@@

Running the test $U_F(j)$ against $Z$ for $j=0,\dots,19$ ($G(P)$ has $p = 4$
states and $G(Z)$ has $q=5$, so $pq-1=19$ is the index of the last test to
be executed  according to Theorem~\ref{th:failurestest}), tests $U_F(0),\dots,
U_F(3)$ are passed by $Z$, but $Z$ fails $U_F(4)$, because after execution of
the trace
\[
s = a.c.c.c, \qquad\text{(note that $G(P)/s = \text{node}\ 3$ according to Fig.~\ref{fig:tga})},
\]
the test $U_F(4)$ offers hitting sets from $\minhits(3) = \{\{b\},\{c\}\}$
in branch (\ref{eq:ufd}). Therefore,  there exists one test execution
where $Z/s$ accepts only $\{b\}$ due to the internal choice (note
from Fig.~\ref{fig:tgZ} that
$G(Z)/s = \text{node}\ 4$), while $U_F(4)/s$
only offers $\{c\}$ in branch (\ref{eq:ufd}) or $\{ a\} = \Sigma - [3]^0$ for
branch (\ref{eq:ufa}).  As a consequence, this execution of
$(Z\parallel[\Sigma] U_F(4))/s$ deadlocks, and the $pass$ event cannot be
produced. Another failing execution arises if $Z/s$ chooses to accept only
$\{c \}$, while $U_F(4)/s$ chooses to accept only $\{a,b\}$. Therefore,
$%\[
(\epass\then \Stop)\not\lessdet_F  (Z\parallel[\Sigma] U_F(4)) \hide\Sigma,
$ %\]
and the test fails. \xbox
\end{example}

@@

With finite complete CSP test suites test suites at hand, it is now possible 
to calculate how many test executions are needed when using them.
Previous work did not consider sufficient conditions for finiteness, so
estimates for the maximal number 
of executions could not be calculated. 
We answer the following questions. (1)~What is
the worst-case bound on the number of test executions to be performed to
verify an SUT with respect to failures refinement, when we use our test
suite? (2)~What is the worst-case bound for traces refinement? (3)~Is it
possible to reduce the maximal length of traces when testing for failures or
traces refinement? We consider the first question~(1) in
Section~\ref{section:complexity:failures}, where we also discuss whether  
it is possible to reduce the number of test executions with a different test
suite.  With the answer to question~(1), question~(2) is a fairly simple
consequence we discuss briefly at the end of 
Section~\ref{section:complexity:failures}. Question~(3) is the subject of
Section~\ref{section:complexity:length}.

@@

last step of the test execution. Therefore, \pagebreak the upper bound

>>>>

Table~\ref{table:operators} lists the CSP operators we use in our examples. A
prefixing operator $e \then P$ defines a process that is ready to engage in
the event $e$, pending agreement of its environment to synchronise. After $e$
occurs, the process behaves as defined by $P$. The environment can be other
processes, in parallel, or the environment of a system as a whole.

\begin{table}\centering
  \begin{tabular}{|l|l|l|}
    \hline
  	$Stop$ & Deadlock process
    \\
  	$a \then P$ & Prefixed operator: initially offers to engage in the event $a$, and then behaves as a process $P$
    \\
  	$g \& P$ & Guarded process: behaves as $P$ if the predicate $g$ is true. Otherwise, it behaves like $STOP$
    \\
  	$P \extchoice Q$ & External choice of $P$ or $Q$ made by the environment
    \\
    $\Extchoice i: I @ P(i)$ & Iterated external choice
    \\
  	$P \intchoice Q$ & Internal choice of $P$ or $Q$
    \\
    $\Intchoice i: I @ P(i)$ & Iterated internal choice
    \\
  	$P \hide X $ & Hiding: behaves like $P$ but with all communications in the set $X$ hidden
    \\
  	$P \parallel[X] Q $ & Generalised parallel: $P$ and $Q$ must synchronise on events that belong to the set $X$
    \\
    \hline
  \end{tabular}
  \caption{{CSP operators:~$P$ and $Q$ are metavariables that stand for processes, $a$ for an event, $g$ for a condition, $X$ for a set of events, and $I$ for a set.}}
  \label{table:operators}
\end{table}

Two forms of choice support branching behaviour. An external choice $P
\extchoice Q$ between processes $P$ and $Q$ offers to the environment the
initial events of $P$ and $Q$. Once a synchronisation takes place, the
process that has offered the event that has occurred is chosen and the other
is discarded. In an internal choice $P \intchoice Q$, the environment does
not have an opportunity to interfere:~the choice is made by the process.
%
\begin{example}\label{example:CSP}
  We consider the processes $P$, $Q$, and $R$ defined below. $P$ is initially
  ready to engage in the event $a$, and then makes an internal choice to
  behave like either $Q$ or $R$.
  \begin{eqnarray*}
  P & = & a \then (Q\intchoice R)
  \\
  Q & = & a \then P \extchoice c \then P
  \\
  R & = & b \then P \extchoice c \then R
  \end{eqnarray*}
  $Q$, for instance, offers to the environment the choice to engage in $a$
  again or $c$. In both cases, afterwards, we have a recursion back to $P$.
  In $R$, if $b$ is chosen, we also have a recursion back to $P$. If $c$ is
  chosen, the recursion is to $R$.
  \xbox
\end{example}
%
Iterated forms $\Extchoice i: I @ P(i)$ and $\Intchoice i: I @ P(i)$ of the
external and internal choice operators define a choice over a collection of
processes $P(i)$. If the index set $I$ is empty, the external choice is the
process $\Stop$, which deadlocks:~does not engage into any event or terminate
successfully. For an external choice $\Extchoice e: A @ e \then P(e),
A\subseteq \Sigma$ over a set $A$ of events, we use the abbreviation $e:A
\then P(e)$. An iterated internal choice is not defined for an empty index
set.

The branches of an external choice can be protected by guards. The process
$$
P(x_1,\dots,x_n) = \text{bexpr}_1(x1,\dots,x_n) \& Q(x_1,\dots,x_n) \extchoice
\text{bexpr}_2(x1,\dots,x_n) \& R(x_1,\dots,x_n)
$$
parametrised over $x_1,\dots,x_n$ branches according to Boolean expressions
$\text{bexpr}_1(x1,\dots,x_n)$ and $\text{bexpr}_2(x1,\dots,x_n)$ over
variables from $\{ x_1,\dots,x_n\}$: if $\text{bexpr}_1(x1,\dots,x_n)$ evaluates to
$\ist$ and $\text{bexpr}_2(x1,\dots,x_n)$ to $\isf$, $P$ behaves like $Q$.
If $\text{bexpr}_1(x1,\dots,x_n)$ evaluates to
$\isf$ and $\text{bexpr}_2(x1,\dots,x_n)$ to $\ist$, $P$ behaves like $R$. If both
Boolean expressions evaluate to $\ist$, $P$ behaves as $Q\extchoice R$. Finally, if
both Boolean expressions evaluate to $\isf$, $P$ behaves like $\Stop$.

@@

Denotational semantic definitions specify, for each operator, how the traces
or failures of the resulting process can be calculated from those of each
operand. For example, for internal choice, $\fails(P\intchoice Q) =
\fails(P)\cup\fails(Q)$; see~\cite[p.~210]{Roscoe:1997:TPC:550448} for a
comprehensive list of these definitions covering $\trc(P)$ and $\fails(P)$.
In this work we use the CSP denotational semantics to justify our results. We
note, however, that the sets of traces and failures of a processes can be
equivalently defined using the operational semantics, as proved
in~\cite{Roscoe2010}. The operational semantics is congruent with the traces
and failures models. This justifies our use of transition graphs as well.

@@

For divergence-free processes, failures refinement, $P \lessdet_F Q$, is
defined by $\fails(Q)\subseteq\fails(P)$. Since refusals are subset-closed,
$P \lessdet_F Q$ implies $(s,\varnothing)\in\fails(P)$ for all traces
$s\in\trc(Q)$. So, for divergence-free processes, failures refinement implies
traces refinement. Therefore, using the conformance relation $conf$ below
%
\begin{equation}\label{eq:conf}
  Q\ conf\ P \defs \forall s\in \trc(P) \cap \trc(Q): \refs(Q/s)
  \subseteq \refs(P/s),
\end{equation}
%
failures refinement can be expressed by $\lessdet_T$ and $conf$ as proven
in~\cite{DBLP:conf/icfem/CavalcantiG07}.
%
\begin{equation}%\label{thm:fref-tref-conf}
\label{eq:failconf}
(P \lessdet_F Q) \iff (P \lessdet_T Q \land Q\ conf\ P)
\end{equation}
%
For finite processes, since refusals are subset-closed, $\refs(P/s)$ can be
constructed from the set of \emph{maximal refusals}.
%
\begin{equation}
\maxrefs(P/s) = \{ R \in\refs(P/s)~|~\forall R'\in \refs(P/s) - \{ R\}: R \not\subseteq R'\}
\end{equation}
%
Conversely, with the maximal refusals $\maxrefs(P/s)$ at hand, we can
reconstruct the refusals in the set $\refs(P/s)$. % as described by
%
\begin{equation}\label{eq:refMaxRef}
\refs(P/s) = \{ R'\in 2^\Sigma~|~\exists R\in \maxrefs(P/s): R'\subseteq R \}.
\end{equation}
%
Deterministic process states $P/s$ have exactly the one maximal refusal
$\Sigma-[P/s]^0$, where $[P/s]^0$ denotes the \emph{initials} of
$P/s$, that is, the events that $P/s$ may engage into. Nondeterministic behaviour in a given process
state is reflected by non-empty intersections between initials and maximal refusals. This
is illustrated by the following example.
%
\begin{example}
\label{ex:nondetdegree} $P = (\Stop \intchoice Q)$ has maximal refusals
$\maxrefs(P) = \{ \Sigma \}$, because $\Stop$ refuses to engage in any event,
and this is carried over to $P$ by the internal choice. However, $P$ is
distinguished from $\Stop$ by its initials, which are defined by $[P]^0 = [
\Stop \intchoice Q]^0 = [Q]^0$. So $P$ may engage nondeterministically in any
initial event of $Q$, but also refuse everything, due to internal selection
of $\Stop$. Assuming an alphabet $\Sigma = \{a,b,c,d\}$, the process
%
$$Q = (e:\{a,b\} \then \Stop) \intchoice (e:\{c,d\} \then \Stop)$$
%
has maximal refusals $\maxrefs(Q) = \{ \{c,d\}, \{a,b\} \}$ and initials
$[Q]^0=\Sigma$. In contrast to $P$, nondeterminism is reflected here by
two maximal refusals. \xbox
\end{example}

@@

% =========================================================================
\paragraph{Graph Products} \label{sec:GP} For proving our main theorems, it is
necessary to consider the \emph{product} of normalised transition graphs. We
need this only for the investigation of corresponding traces in reference
processes and processes for SUTs. So, the labelling of nodes with maximal
refusals or minimal acceptances is disregarded in the product construction.
We consider two normalised transition graphs over the same alphabet $\Sigma$.
\[
G_i = ( N_i, \ii n_i, \Sigma, t_i : N_i\times\Sigma \pfun N_i, r_i : N_i \fun \mathbb{P}\mathbb{P}(\Sigma)),\qquad i = 1,2,
\]
Their product is defined by
%
\begin{eqnarray}
G_1\times G_2 & = & (N_1\times N_2,(\ii n_1,\ii n_2), t:(N_1\times N_2)\times\Sigma\pfun (N_1\times N_2))
\\
\dom~t & = & \{ ((n_1,n_2),e)\in (N_1\times N_2)\times\Sigma | (n_1,e)\in\dom~t_1\wedge (n_2,e) \in\dom~t_2 \}
\\
t((n_1,n_2),e) & = & (t_1(n_1,e),t_2(n_2,e))\ \text{for $((n_1,n_2),e)\in\dom~t$}
\end{eqnarray}
%
The following lemma is used later in the proof of our main theorem.
%
\begin{lemma}\label{lemma:reachproduc}
If $G_1$ has $p$ states and $G_2$ has $q$ states, then every reachable state
$(n_1,n_2)$ of the product graph $G_1\times G_2$ can be reached by a trace
%%%$s\in\trc(G_1)\cap \trc(G_2)$  we don't need this
of maximal length $(pq-1)$.
\end{lemma}
\begin{proof}
The product graph $G_1\times G_2$ has at most $pq$ states. The empty trace $\varepsilon$
reaches its initial state $(\ii n_1,\ii n_2)$. Applying Lemma~\ref{lemma:extendV}
$(pq-1)$ times with $V=\{\varepsilon \}$ implies that $G_1\times G_2$ reaches
all of its reachable states (there are at most $pq$ of them) under
$V' = V \cup V.\Sigma\cup\dots \cup V.\Sigma^{(pq-1)}$. The maximal length of traces in
$V'$ is $(pq-1)$.
\xbox
\end{proof}
%
This concludes our presentation of CSP and of results regarding its
semantics that are used in the next section.

@@

\paragraph{Definition} The main idea of the underlying test strategy for failures
refinement is based on solving a \emph{hitting set problem}. Given a finite
collection of finite sets ${\cal C} = \{ A_1,\dots,A_n\}$, such that each
$A_i$ is a subset of a universe $\Sigma$, a \emph{hitting set}
$H\subseteq\Sigma$ is defined as a set satisfying the following property.
%
\begin{equation}
  \label{eq:hit}
  \forall A\in {\cal C}: H\cap A \neq\varnothing.
\end{equation}
%
A \emph{minimal hitting set} is a hitting set that cannot be further reduced
without losing the characteristic property (\ref{eq:hit}). By $\minhits({\cal
C})$ we denote the collection of minimal hitting sets for a collection ${\cal
C}$. For the pathological case where $C$ contains an empty set,
$\minhits({\cal C})$ is also empty. The problem of determining minimal hitting sets is %%% \cite{Book1975-BOOKRM,5533149}
NP-hard~\cite{5533149}. We see below, however, that using minimal hitting
sets, we can reduce the effort of testing for failures refinement from a
factor of $2^{|\Sigma|}$ to a factor that equals the number of minimal
hitting sets.

% -----------------------------------------------------------------------------
\paragraph{Minimal Hitting Sets of Minimal Acceptances} In this article, we are
interested in the minimal hitting sets of minimal acceptances; for these, the
abbreviated notation
$%\[ \
\minhits(P/s) = \minhits(\minaccs(P/s))
$ %\]
is used. The minimal hitting sets
of minimal acceptances may be alternatively characterised by means of the
failures of a process as is done in~\cite{DBLP:conf/icfem/CavalcantiG07}. To
this end, in~\cite{DBLP:conf/icfem/CavalcantiG07}, the authors define, for
any collection ${\cal C}\subseteq 2^\Sigma$ of subsets from $\Sigma$
\begin{equation}\label{eq:minsub}
\minsub({\cal C}) = \{ A\in {\cal C}~|~\forall B\in {\cal C}:
B\subseteq A\Rightarrow B = A \}.
\end{equation}
The collection $\minsub({\cal C})$ contains all those sets of ${\cal C}$ that
are not true supersets of other members of ${\cal C}$. With this definition,
the relation between failures and minimal hitting sets of minimal acceptances
is established as follows.
%
\begin{lemma}
\label{lemma:failureshittingsets}
For any trace $s$ of CSP process $P$, define
$%\[
{\cal A}_s = \{ A\subseteq \Sigma~|~(s,A)\not\in\fails(P)  \}.
$ %\]
Then
$ %\[
\minhits(P/s) = \minsub({\cal A}_s)
$ %\]
for all traces $s$ of $P$.
\end{lemma}
\begin{proof}
We derive
\begin{argue}
(s,A)\not\in \fails(P)
\\
\Leftrightarrow A\not\in Ref(P/s)
\\
\Leftrightarrow  \forall R\in maxRef(P/s): A\not\subseteq R
\\
\Leftrightarrow   \forall B\in minAcc(P/s): A\not\subseteq (\Sigma -
B)& this follows from (\ref{eq:accref})
\\
\Leftrightarrow  \forall B\in minAcc(P/s): A\cap B\neq \varnothing
\\
\Leftrightarrow  A \text{ is a (not necessarily minimal)}
              \text{ hitting set of } minAcc(P/s)
\end{argue}
%
This derivation is valid  for arbitrary $A$ satisfying $(s,A)\not\in {\text{failures}}(P)$.
Specialising it to {\it minimal} sets $A$ satisfying  $(s,A)\not\in {\text{failures}}(P)$
proves the statement of the lemma.
\xbox
\end{proof}

@@

To prove ``$(2) \Rightarrow (1)$'', we assume that (2) holds, but that $P\
conf\ Q$ does {\it not} hold. According to Lemma~\ref{lemma:tgtrcref},
(\ref{eq:failrefb}), there exists $s\in\trc(P)\cap \trc(Q)$ that satisfies
the following property.
\[
\exists A_Q\in\minaccs(G(Q)/s): \forall A_P\in\minaccs(G(P)/s):
A_P\not\subseteq A_Q \qquad (*)
\]
Let $A$ be such an acceptance set $A_Q$ fulfilling (*).
Define
$%\[
\overline H = \bigcup\{ A_P \setminus A~|~A_P\in\minaccs(G(P)/s) \}.
$ %\]
Since $A_P \setminus A \neq\varnothing$ for all $A_P$ because of (*),
$\overline H$ is a hitting set of $\minaccs(G(P)/s)$ which has an  empty
intersection with $A$. Minimising $\overline H$ yields   a minimal hitting
set $H\in \minhits(P/s)$ which is {\it not} a hitting set of
$\minaccs(G(Q)/s)$, a contradiction to Assumption~2.
This completes the proof of the lemma.
\xbox
\end{proof}
%
We note that $\minaccs(P) = \{ \varnothing \}$ if $P = Q\intchoice \Stop$.
Since $\Stop$ accepts nothing, its minimal acceptance is $\varnothing$, and
this carries over to $Q\intchoice \Stop$.  From (\ref{eq:failrefb}) we
conclude that $\varnothing\in\minaccs(P)$ implies $\minaccs(P) = \{
\varnothing \}$. This clarifies that $\minhits(P/s)$ is empty if, and only
if, $\minaccs(P) = \{ \varnothing \}$. The proof of Lemma~\ref{lemma:hseta}
covers the situations where $\minaccs(P/s) = \{ \varnothing \}$ and so
$\minhits(P/s) = \varnothing$. Trivially,
\begin{equation}
\label{eq:minhitminaccempty}
\minaccs(P/s) = \{ \varnothing \} \iff \minhits(P/s) = \varnothing
\end{equation}
holds.

@@

It is assumed that the events $\efail$ and $\epass$, indicating verdicts FAIL
and PASS for the test execution, are not included in $\Sigma$. Since we
assume that $Q$ is free of livelocks, the events $\efail$ or $\epass$ always
become visible, if they are the only events $U_F(j)/s$ is ready to engage in:
if $U_F(j)/s$ can only produce $\epass$ or $\efail$, the occurrence of these
events can never be blocked due to a livelock, since the only
  process executing concurrently with  $U_F(j)$  is livelock-free  $Q$.

@@

Running the test $U_F(j)$ against $Z$ for $j=0,\dots,19$ ($G(P)$ has $p = 4$
states and $G(Z)$ has $q=5$, so $pq-1=19$ is the index of the last test to be
executed  according to Theorem~\ref{th:failurestest}), tests $U_F(0),\dots,
U_F(3)$ are passed by $Z$, but $Z$ fails $U_F(4)$, because after execution of
the following trace.
\[
s = a.c.c.c \qquad\text{(Note that $G(P)/s = \text{node}\ 3$ according to Fig.~\ref{fig:tga}.)}
\]
The test $U_F(4)$ offers hitting sets from $\minhits(3) = \{\{b\},\{c\}\}$ in
branch (\ref{eq:ufd}). Therefore,  there exists one test execution where
$Z/s$ accepts only $\{b\}$ due to the internal choice (note from
Fig.~\ref{fig:tgZ} that $G(Z)/s = \text{node}\ 4$), while $U_F(4)/s$ only
offers $\{c\}$ in branch (\ref{eq:ufd}) or $\{ a\} = \Sigma - [3]^0$ for
branch (\ref{eq:ufa}).  As a consequence, this execution of
$(Z\parallel[\Sigma] U_F(4))/s$ deadlocks, and the $pass$ event cannot be
produced. Another failing execution arises if $Z/s$ chooses to accept only
$\{c \}$, while $U_F(4)/s$ chooses to accept only $\{a,b\}$. Therefore,
$%\[
(\epass\then \Stop)\not\lessdet_F  (Z\parallel[\Sigma] U_F(4)) \hide\Sigma,
$ %\]
and the test fails. \xbox
\end{example}

@@

With finite complete CSP test suites, it is possible to calculate how
many test executions are needed when using them. Previous work did not
consider sufficient conditions for finiteness, so estimates for the maximal
number of executions could not be calculated. We answer the following
questions. (1)~What is the worst-case bound on the number of test executions
to verify an SUT with respect to failures refinement, when we
use our test suite? (2)~What is the worst-case bound for traces refinement?
(3)~Is it possible to reduce the maximal length of traces when testing for
failures or traces refinement? We consider question~(1) in
Section~\ref{section:complexity:failures}, where we also discuss whether it
is possible to reduce the number of test executions with a different test
suite.  With the answer to question~(1), question~(2) is a simple consequence
we discuss briefly at the end of Section~\ref{section:complexity:failures}.
Question~(3) is the subject of Section~\ref{section:complexity:length}.

@@

last step of the test execution. Therefore, the upper bound
